Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-06-05 Manuscript/proto-manuscript.tex

why did I choose Lipid 14 -- good ratio of $\alpha / \beta$ response~\cite{catte16}.

@@

Detailed and robust structural information of lipid bilayers can
be reached with C-H bond order parameter from NMR and form factors from scattering
experiments. The former can be used to validate structures sampled by individual
molecules and latter to validate overall bilayer properties, lateral density
(area per molecule) and thickness \cite{ollila16}. The structural quality of
new ECC lipid model against scattering form factors and NMR order parameters
is evaluated in Fig. \ref{simVSexpNOions}
\todo{Add acyl chain order parameters, POPC chemical structure and fix experimental alpha order parameter value}
\todo{Why original Lipid14 models seems to deviate from experiments, in contrast to the original publication?}
\begin{figure}[]
  \centering
  \includegraphics[height=8.5cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
%  \caption{\label{fig:form-f}
%    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 and ECCLipids17 models.  }
%\end{figure}
%
%\begin{figure}[]
%  \centering
  \includegraphics[width=9.0cm]{../Fig/ipython_nb/Headgr_OPs_exp-L14-ECCL17.png}
  \caption{\label{simVSexpNOions}
    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 and ECCLipids17 models. 
    Headgroup and glycerol backbone order parameters from simulations with Lipid14 \cite{dickson14} and EECLipid17 models
    compared with experimental order parameters from \cite{ferreira13}.}
\end{figure}
Area per molecules extracted from MD simulations and SPD model fitted to scattering data \todo{check that this the case for the used values}
are shown in Table \ref{apls}
\todo{finalize figure and the discussion}
\begin{table}
  \caption{Area per lipid from different models for POPC without ions\label{apls} }
  \begin{tabular}{c c c}
    model          & A (Å$^2$)   & Temperature [K] \\
    \hline
    Lipid14 (literature)  & 65.6$\pm$ 0.5  &  303 \\
    Lipid14ecc0.80+sigma0.875 &        &  313    \\
    GMX small patch           & 64.9   &         \\
    GMX 4xbig patch           & 65.5   &         \\
    oMM small patch           & 63.65  &         \\
    oMM 4xbig patch           & 63.7   &         \\
    \hline
    experiment \cite{Jambeck2012}\todoii{REF}{put original references, not Slipids param. paper.}  & 62.7  &  293    \\
    experiment  & 64.3  &  303    \\
    experiment  & 67.3  &  323    \\
    experiment  & 68.1  &  333    \\
    experiment POPE  & 56.6 &  303    \\
    \hline
  \end{tabular}
\end{table}

%Preserves other relevant properties: APL, Order Parameters (OPs of headgroup and tails), 
%diffusion? \todoi{diffusion can be even improved -- and doesn't have to. We may as well omit it.}, 
%elasticity? \todoi{check elastic properties -- do they change, improve, worsen?? -- likely not crucial for the main message of the paper.} 
%-- however, it is still garbage in-garbage out
%The glycerol backbone and headgroup order parameters are plotted in
%Fig \ref{HGopsNEWmodels.eps} for reference models and EEC models. The difference,
%i.e. the effect of scaling on these parameters, is shown in Fig. \ref{HGopsNEWmodelsCHANGE}.

%\textbf{Required figures}

%- Figure showing the system: POPC, membrane, atoms involved in headgroup OPs.\todoi{This is probably sufficient as the last thing we would do -- and will probably provide a good basis for TOC graphic.}

%- Figure comparing order parameters (Fig.~\ref{HGopsNEWmodels}) and form factor \todoi{The original L14 form-f are worse than those in the L14 paper -- I'd better double check it and make sure I use the same settings and version for all simulations consistently for a good comparison.}between original Lipid14, new model and experiments,

Ion binding in lipid bilayers can be accurately measured and compared
between experiments and simulations by using lipid headgroup order
parameters and ''electrometer concept'' introduced by Seelig et al. \cite{seelig87,catte16}.
This is based on the experimental observations that order parameters
for $\alpha$ and $\beta$ carbons are proportional to the amount of bound charge
in lipid bilayer, which was rationalized as a carge induced tilt
of headgroup dipole \cite{seelig87}. Later analysis including signs of the order
parameters showed that the order parameters are actually decreasing with bound
positive charge and {\it vice versa} for negative charge \cite{ollila16,catte16}.

Order parameter changes as a function of bound charge are shown in
Fig. \ref{OrderParameterCHANGESnewMODELS}. Approximately linear
decrease of headgroup order parameters is observed in simulations and
experiments with cationic surfactant. Quantitative comparison reveals
that in original Lipid14 model the proportionality constant is
overestimated, thus the headgroup is too sensitive to the bound cations.
However, the new ECC lipid model shows better agrement, althoght
$\beta$ responce is slightly underestimated.

{\it What affects the headgroup OP response?: The response of the headgroup order parameters is improved
both through diminished affinity towards the cations and through lower headgroup response to the bound charge
(from simulations with \ce{DHMDMAB} surfactant). 
The responses of headgroup OPs are shown in Fig. \ref{OrderParameterCHANGESnewMODELS}. }

%Accurate binding of \ce{Na^+} and \ce{Ca^{2+}} cations. 
%Accurate electrometer response to \ce{Na^+}, \ce{Ca^{2+}} and \ce{DHMDMAB} cationic surfactant. 
%- Figure showing order parameters as a function of \ce{NaCl} and \ce{CaCl_2} concentrations~

@@

\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Fig/OrdParChanges_NaCl_CaCl2_surf.eps}
  \caption{\label{OrderParameterCHANGESnewMODELS}
    Headgroup order parameter changes as a function of NaCl, CaCl$_2$ concentration and
    cationic surfactant (dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$)
    from simulations and experiments (DPPC \cite{akutsu81}, POPC \cite{altenbach84}, surfactant \cite{scherer89}).
    Simulations with Lipid14 and Åqvist ion model from \cite{catte16,lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC1000mMCaClfiles}.
  }
  \todo{Lipid14/Åqvist data to be added from https://github.com/NMRLipids/lipid\_ionINTERACTION/blob/master/Data/POPC/CaCl/LIPID14/LIPID14caclCONSchange.dat}
\end{figure*}

Ion density profiles between different simulation models are compared in Fig. \ref{fig:cacl-dens}.
Density profiles from simulations with original Lipid14 and Åqvist ions and new ECC lipid model with
scaled ions are not much different, thus the overestimated order parameter decrease is explained
by the overestimated headgroup response on bound charge in original lipid model (see Fig. \ref{OrderParameterCHANGESnewMODELS}). 
Introducing scaling in only on ion models increase the ion binding which is also seen in order parameter
changes in Fig. \ref{OrderParameterCHANGESnewMODELS} \todo{Final check to be done when the data from Åqvist model is there}.
The increased binding can be probably explained by the changes in the size of the ion due to scaling \todo{more detailed explanation?}.
Using ECC approach also in lipid headgroups decreases the ion binding and also brings headgroup response
on bound charge close to experiments in Fig. \ref{OrderParameterCHANGESnewMODELS}. Alltogether,
the response of headgroup order parameters as a function of CaCl$_2$ concentration is in agreement with
experiments in new ECC lipid model simulated with scaled CaCl$_2$ model. This is a significant improvement
on previous modela available for lipid--ion interactions studies \cite{catte16} and indicates that
the new model can be used for detailed studies of lipid-ion interactions.

\todo{Analyze stoichiometry for \ce{Na^+}, \ce{Ca^{2+}}, their interaction energies with the lipid membrane, etc, and
finalize the discussion after these results.}

@@

>>>>

why did I choose Lipid 14 -- good ratio of $\alpha / \beta$ response~\cite{catte16} but not perfect.

Detailed and robust structural information of lipid bilayers can
be reached with C-H bond order parameter from NMR experiments,
which tell on the lipid structure sampled by each individual molecule.
On the other hand, form factors from scattering experiments validate overall bilayer properties, 
lateral density (area per molecule) and thickness \cite{ollila16}. 
The structural quality of
new ECC lipid model is evaluated against scattering form factors and NMR order parameters
in Fig. \ref{simVSexpNOions}
\todo{Add acyl chain order parameters, POPC chemical structure}
\todo{Why original Lipid14 model seems to deviate from experiments, in contrast to the original publication? Different simulation setup used?}
\begin{figure}[]
  \centering
  \includegraphics[height=8.5cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
%  \caption{\label{fig:form-f}
%    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 and ECCLipids17 models.  }
%\end{figure}
%
%\begin{figure}[]
%  \centering
  \includegraphics[width=9.0cm]{../Fig/ipython_nb/Headgr_OPs_exp-L14-ECCL17.png}
  \caption{\label{simVSexpNOions}
    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 and ECCLipids17 models. 
    Headgroup and glycerol backbone order parameters from simulations with Lipid14 \cite{dickson14} and EECLipid17 models
    compared with experimental order parameters from \cite{ferreira13}.}
\end{figure}
Area per molecules extracted from MD simulations and SPD model fitted to scattering data \todo{check that this the case for the used values}
are shown in Table \ref{apls}
\todo{finalize figure (NMR headgr. OPs + SAXS, continue in the discussion.}
\begin{table}
  \caption{Area per lipid from different models for POPC without ions\label{apls} }
  \begin{tabular}{c c c}
    model          & A (Å$^2$)   & Temperature [K] \\
    \hline
    Lipid14 (literature)  & 65.6$\pm$ 0.5  &  303 \\
    Lipid14ecc0.80+sigma0.875 &        &  313    \\
    GMX small patch           & 64.9   &         \\
    GMX 4xbig patch           & 65.5   &         \\
    oMM small patch           & 63.65  &         \\
    oMM 4xbig patch           & 63.7   &         \\
    \hline
    experiment \cite{Jambeck2012}\todoii{REF}{put original references, not Slipids param. paper.}  & 62.7  &  293    \\
    experiment  & 64.3  &  303    \\
    experiment  & 67.3  &  323    \\
    experiment  & 68.1  &  333    \\
    experiment POPE  & 56.6 &  303    \\
    \hline
  \end{tabular}
\end{table}

Ion binding in lipid bilayers can be accurately measured and compared
between experiments and simulations by using lipid headgroup order
parameters and ''electrometer concept'' introduced by Seelig et al. \cite{seelig87,catte16}.
This is based on the experimental observations that order parameters
for $\alpha$ and $\beta$ carbons are proportional to the amount of bound charge
in lipid bilayer, which was rationalized as a charge induced tilt
of the headgroup dipole \cite{seelig87}. Later analysis including signs of the order
parameters showed that the order parameters are actually decreasing with bound
positive charge and {\it vice versa} for negative charge \cite{ollila16,catte16}.

@@

Order parameter changes as a function of bound charge are shown in
Fig. \ref{OrderParameterCHANGESnewMODELS}. 
Approximately linear decrease of headgroup order parameters is observed both in simulations and experiments with the cationic surfactant. 
The use of a cationic surfactant has the benefit that it not a subject of partitioning between water and lipid phase. 
This means that we know the exact amount of bound charge in the membrane both in experiment and simulaiton. 
Quantitative comparison reveals that in the original Lipid14 model the response is overestimated for both order parameters, $\alpha$ and $\beta$. 
This suggests that the observed overestimated response in \cite{catte16} 
is at least in part due to the high sensitivity of the headgroup order parameter response, 
not only due to overestimated binding of cations. 
Secondly, the $S^\alpha/S^\beta$ ratio of the response is found to be slightly larger than in experiments (VALexp vs VALsim\todoi{Add values of $S^\alpha/S^\beta$ response for sim and experiment}).
This property is kept also by the newly derived model, ECClipid, and, hence, 
it shows a perfect agreement in the response of the order parameter $S^\alpha$ 
but underestimates slightly the response of $/S^\beta$.

@@

\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Fig/OrdParChanges_NaCl_CaCl2_surf.eps}
  \caption{\label{OrderParameterCHANGESnewMODELS}
    Headgroup order parameter changes as a function of NaCl, CaCl$_2$ concentration and
    cationic surfactant (dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$)
    from simulations and experiments (DPPC \cite{akutsu81}, POPC \cite{altenbach84}, surfactant \cite{scherer89}).
    Simulations with Lipid14 and Åqvist ion model from \cite{catte16,lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC1000mMCaClfiles}.
  }
  \todoii{Add Lipid14-Aquist data.}{Lipid14/Åqvist data to be added from https://github.com/NMRLipids/lipid\_ionINTERACTION/blob/master/Data/POPC/CaCl/LIPID14/LIPID14caclCONSchange.dat . Joe: I'm not sure what is meant -- I think Aquist data are actually plotted there. I recently changed it to L14+Dang ions (both data and plot). I think just one ion model is sufficient (and ECC-ions are based on Dang model).}
\end{figure*}

Ion density profiles between different simulation models are compared in Fig. \ref{fig:cacl-dens}.
Density profiles from simulations with original Lipid14 and Dang ions \cite{smith94,chang1999,dang2006} show a pronounced peak in the position of the phosphate moieties of POPC. 
The use of a ECC-ion model~\cite{kohagen14,kohagen16} along with original Lipid14 does not significantly change it. 
The new ECClipid model with scaled ions exhibits on the other hand smaller density in this region suggesting overall weaker binding of cations (Fig.~\ref{OrderParameterCHANGESnewMODELS}). 
This demonstrates how importnat it is to account for electronic polarizability also in molecules with zero total charge. 
Alltogether,
the response of headgroup order parameters as a function of CaCl$_2$ concentration is in agreement with
experiments in the new ECClipid model simulated with ECC model of CaCl$_2$. This is a significant improvement
over previous models available for lipid--ion interactions studies \cite{catte16} and indicates that
the new model can be used for detailed studies of lipid-ion interactions.

@@

For example, there is a debate in the literature on the stoichiometry of \ce{Ca^{2+}} binding to POPC~\cite{Altenbach84} \todoii{REFs}{Add references on Ca2+:POPC stoichiometry}. 
In line with the early experimental finding~\cite{Altenbach84}, we find that our data fit well a ternary complex model, which assumes 2 POPC molecules per 1 \ce{Ca^{2+}}. 
In contrast to this experiment, however, we possess also a complete atomistic detail of this phenomenon.
Hence we can support this empirical finding with a direct observation of transient complexes of \ce{Ca^{2+}} and POPC.
The simulation suggests that the stoichiometry 2~POPC:1~\ce{Ca^{2+}} is the most common complex. 
In addition, there also is a non-negligable probability of one \ce{Ca^{2+}}  cross-bridging three POPC molecules. 
This cannot be observed in experiments without atomistic detail as it makes only a small perturabation to the binding isotherm that assume 2~POPC per 1~\ce{Ca^{2+}}. 
\todo{Analyze stoichiometry for \ce{Na^+}, \ce{Ca^{2+}}, their interaction energies with the lipid membrane, etc, and finalize the discussion after these results.}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm,angle=0]{../Fig/CA_binding.eps}
  \caption{\label{fig:cacl-bind}
    Binding isotherm assuming stoichiometry of 2~POPC:1~\ce{Ca^{2+}} as used in \cite{Altenbach84} fits the simulation data nicely.}
\end{figure}
